---
title: "Analyzing Incucyte data with tidycyte"
description: >
  The tidycyte library enables efficient and simple manipulation of Incucyte raw data matrices
  using dplyr, purrr, and other features of the tidyverse. The result is a simple, streamlined
  pipeline that parses Incucyte data matrices and enables normalization and arithmetic
  operations with only a few key intuitive commands. The tidycyte workflow is designed to
  prepare data sets for visualization using ggplot2.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing Incucyte data with tidycyte}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
set.seed(1014)
```

## Introduction

The philosophy of tidy data is at the core of tidycyte, and as a result,
this code makes heavy use of dplyr, purrr, and tidyr.
Generally, the tidycyte pipeline is designed to efficiently return tidy
data frames, meaning that each row is an individual
observation, and each column represents a unique value associated with the
observation.

As tidycyte is built on dplyr, almost all operations rely on
data masking (see "Additional information" below). This concept
permits accessing column names directly as though they are variables,
rather than referencing the data frame at each mention.

Additionally, the pipeline is amenable to the use of the magrittr pipe
(`%>%`), which streamlines the overall pipeline to focus on the flow of
data operations.

This vignette will provide the minimum knowledge you need to effectively
use tidycyte.

```{r, eval=F, echo=T}
library(tidycyte)
```

## Building the metadata table

The metadata table is a key part of the workflow that specifies how the
samples relate to each other for the purposes of normalization, as well as
time points that are to be used for normalization. It can be created as a
data frame, read in using `read.table()`, or, perhaps
the easiest way to build the metadata table is to use the
`tribble()` function as shown below:

```{r, eval=F, echo=T}
metadata <- tribble(
  ~metric,       ~filename,        ~metric_norm, ~time_norm, ~ymin, ~ymax,
  "Confluence",  "confluence.txt", NA,           NA,         0,     75,
  "h2-3",        "green.txt",      "Confluence", 0,          0,     6,
  "mAzalea",     "red.txt",        "Confluence", 0,          0,     5)
```

At the minimum, `metadata` needs `metric` and `filename` columns to read
the Incucyte matrix files. To perform normalization of the data,
`metric_norm` and `time_norm` will also need to be
provided as columns. Extra columns do not present a problem and the table
can contain as many extra columns as desired.

## A typical tidycyte pipeline

The exact pipeline for each tidycyte analysis will depend on the user's
preferences and the desired analysis. Below is a sample workflow:

```{r, eval = F, echo = T}
df <- parse_incucyte_matrices(metadata) %>%
  sum_across_images() %>%
  normalize_by_metric(metadata) %>%
  normalize_by_time(metadata)
```

Except for the `parse_incucyte_matrices()` call,
which must come first, the order of the commands can be modified by the
user. The above order is highly recommended; however, if desired, time
normalization could be before metric normalization, or averaging images
could happen after all of the normalization. Elements of the pipeline can
also be ommitted if desired.

The key idea behind the use of the magrittr pipe (`%>%`) is to simplify and
streamline the pipeline steps, allowing the user to focus on data
processing, rather than coding syntax.

## Metric manipulation operations

Several functions are provided for the manipulation of metrics. These are
briefly listed below, and users are encouraged to read their documentation
and sample for themselves how the functions modify the data. Example usage
for each function is provided within the respective help documents.

*  `sum_across_images()` sums metrics when multiple images are taken per well.
*  `average_across_images()` averages metrics when multiple images are taken per well.
*  `replace_na_value_in_metric()` replaces NA values in specified metrics.
*  `replace_minimum_value_in_metric()` replaces values below a minimum threshold in specified metrics.
*  `replace_maximum_value_in_metric()` replaces values above a maximum threshold in specified metrics.
*  `derive_metric()` is one of the most versatile functions that permits mathematical operations on metrics.
*  `scale_metric()` scales a specified matric to span a new range (e.g. min to max scaling).
*  `metrics_to_fractions()` converts object counts among specified metrics to fractions.
*  `summary_stats()` calculates summary statistics, including the standard error across measurements.

## Filtering and specifying plotting order

The main data frame can be filtered using the `filter()`
command from `dplyr`. If desired, multiple filtering steps can occur in
one command:

```{r, eval = F, echo = T}
df <- df %>% filter(treatment %in% treatments_to_plot, elapsed %in% elapsed_to_plot)
```

Additionally, factor levels can be specified for controlling plotting
order using `mutate()` from `dplyr`:

```{r, eval = F, echo = T}
df <- df %>% mutate(treatment = factor(treatment, levels=treatments_to_plot))
```

## Sample dataset

The library comes with a sample dataset that corresponds to measurement
of confluency, as well as green and red fluorescent cell cycle markers
(FUCCI). The sample dataset, `cyclesync`, is automatically available through
"lazy" data loading, and it can be accessed by name at any time. This dataset
can be helpful for learning how the pipeline works, and aid in the development
of new pipelines or functions. It corresponds to the output of
`parse_incucyte_matrices()`.

To access the dataset, just treat it like you would any other variable. Its
corresponding metadata table is the example `metadata` table provided in the
"Building the metadata table" section above.

```{r, eval = F, echo = T}
df <- cyclesync
```

## Summary statistics

If desired, rather than plotting individual measurements, summary
statistics can be calculated for plotting of means with error bars:

```{r, eval = F, echo = T}
dfs <- df %>% summary_stats(value,elapsed,treatment,cell,metric,.ci = 0.95)
```

## A complete pipeline

If all variables are known in advance, the entirety of a tidycyte workflow
can be accomplished in a single pipeline:

```{r, eval = F, echo = T}
df <- parse_incucyte_matrices(metadata) %>%
  sum_across_images() %>%
  normalize_by_metric(metadata) %>%
  normalize_by_time(metadata) %>%
  filter(treatment %in% treatments_to_plot, elapsed %in% elapsed_to_plot) %>%
  mutate(treatment = factor(treatment, levels=treatments_to_plot)) %>%
  summary_stats(value,elapsed,treatment,group,metric,.ci = 0.95)
```

At this stage, `df` can be used directly in `ggplot()`.

## Additional information

Many syntax issues are covered in the resources below. These can be accessed
by typing these commands in the console.

*  `help(package="tidycyte")`
*  `vignette("dplyr")`
*  `?dplyr::dplyr_data_masking`

------
The tidycyte library is Copyright 2021 by H. Courtney Hodges.
